High performance computing for global optimization problems

### 2017-11-14 2018/ITNT/ITNT2018.tex
version for submission

\address{Lobachevsky State University of Nizhni Novgorod, Gagarin avenue 23, 
Nizhni Novgorod, Russia, 603950}

@@

\section{Introduction}
In the present paper, the constrained multiextremal optimization problems and 
the parallel methods for solving these ones are considered. The fact that the 
global extremum is an integral characteristic of a problem is an important 
feature of the multiextremal problems. Thus, its finding is related to the 
construction of a coverage of the search domain and the computing the 
optimized function values at all points of this coverage. The dimensionality 
affects the difficulty of solving the problems of considered class crucially: 
the computational costs grow with increasing this one exponentially, 
therefore, for solving such problems, the methods are required, which 
generate an essentially nonuniform grid in the search domain, denser in the 
nearness of the global minimum and more sparse far away from this one. In the 
present paper, the results of investigation of the scaling of the parallel 
index global optimization algorithm developed in Lobachevsky State 
University of Nizhni Novgorod \cite{Strongin2000,Strongin2013} and implemented in the Globalizer 
system [NACO] are presented.

@@

In the algorithms developed in the present paper the objective function and constraints are assumed to be Lipschitzian that is typical for many other approaches to the development of the parallel optimization algorithms (see, for example, \cite{Jones,Zilinskas,Evtushenko}). This assumption is natural for many applied problems since the relative variations of the functions generally cannot exceed a certain threshold determined by the bounded energy of changes occurring in the system under study. Also, a standard assumption is that the execution of a single trial (the computing of the values of the objective function and constraints at a point in the search domain) is a time-consuming operation since it 
utilized the results of numerical modeling.

An assumption on different costs of computing the problem functions subject to 
various components of the vector of parameters is a novel element investigated in the present work. 
It is assumed that it is possible to select the ``difficult'' and ``easy'' parts 
of the functions connected with each other. Then, the solving of the 
``difficult'' part of the problem (for which the execution of the computation-
costly algorithms is required to conduct the trials) can be performed on the 
CPU using an efficient index global optimization algorithm. The ``easy'' part 
of the problem (the execution of which doesn’t require the complex algorithms 
and can be transferred to an accelerator easily) can be solved on a GPU using 
the uniform grid technique. Obviously, in this case a major portion of the trials 
will be performed on the GPUs, and the minor one -- on the CPU. However, 
because of the difference in the difficulties of the subproblems solved on 
different units, a speedup of the algorithm as a whole can be expected. 
The proposed approach has been implemented in Globalizer parallel software 
system for solving the global optimization problem developed in Lobachevsky 
State University of Nizhni Novgorod.

@@

The use of Peano curve $y(x)$ 
\begin{equation}
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq 
N\right\}=\left\{y(x):0\leq x \leq 1 \right\}
\end{equation}
unambiguously mapping the interval of real axis $[0,1]$ onto a $N$-
dimensional cube is the first of the dimensionality reduction methods 
considered.  To implement this method of dimensionality reduction a 
numerically constructed curve (\textit{evolvent}) is used. The evolvent is 
$2^{-m}$ accurate approximation of the theoretical Peano curve in $L_\infty$ 
metric, where $m$ is an evolvent construction parameter. Problems of 
numerical construction of the evolvents and the corresponding theory are 
considered in detail in \cite{Strongin2000}.

@@

Nested optimization scheme is based on relation (see \cite{Strongin2013})
\begin{equation}\label{nested}
\min_{y \in D}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq 
m\right\}}= \min_{u_1\in D_1}\min_{u_2\in D_2}...\min_{u_M\in D_M 
}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq m\right\}},
\end{equation}
which allows replacing the solving of multidimensional problem 
(\ref{problem}) by solving a family of subproblems related to each other 
recursively.
Here we consider vector $y$ as a vector of block variables
\begin{equation}
y=(y_1,y_2,...,y_N)=(u_1,u_2,...,u_M),
\end{equation}
where the $i$-th block variable $u_i$ is a vector of $N_i$ components of 
vector $y$, taken serially i.e. $u_1=(y_1,y_2,...,y_{N_1})$, 
$u_2=(y_{N_1+1},y_{N_1+2},...,y_{N_1+N_2})$,..., $u_M=(y_{N-N_M+1},y_{N-
N_M+2},...,y_{N})$, at that $N_1+N_2+...+N_M=N$. 
The subdomains $D_i, 1 \leq i \leq M$, are projections of initial search 
domain $D$ onto the subspaces corresponding to the variables $u_i, 1 \leq i 
\leq M$.

@@

The organization of parallel computing with the use of the recursive 
optimization scheme has been considered in details for the shared/distributed 
memory as well as for the accelerators in \cite{Sysoyev,BarkalovLebedev}. However, in these works the 
problems were considered, in which the time of the trial execution didn’t 
depend on the trial point. Here, the problem is considered assuming that in 
the function $\varphi(y_1,...,y_N)$ it is possible to select more difficult 
part $f(y_1,...,y_N)$ (depending on all problem parameters) and a simpler 
part $g(y_s,...,y_N )$ (depending on a part of the parameters only), for 
example, $\varphi(y)= f(y)g(y)$.

The difficult part of the function implies performing some  
computations related to the numerical simulation, which can be performed on the 
CPUs only. The simple part doesn’t imply the complex computations and can be 
computed on an accelerator, for example, GPU. For solving a problem with such 
a structure, one can apply the parallel recursive optimization scheme with 
the use the CPUs at the upper nesting level and the GPUs at the lower one.

@@

The recursive scheme of solving the global optimization problems has been 
implemented in the Globzlizer solver developed in Lobachevsky State 
University of Nizhni Novgorod. The global search methods and various 
dimensionality reduction schemes make the algorithmic base for the 
Globalizer. The numerical experiments, the results of which are presented in 
\cite{BarkalovGergelLebedev,BarkalovGergel} demonstrate these methods, at least, are not worse than the well known 
methods applied for similar purposes, and even overcome these ones with 
respect to some parameters.

@@

The level lines of the two-dimensional subproblems based on the functions 
$f(y_1,y_2)$ and $g(y_3,y_4)$ are shown in figure~1. The feasible domains are highlighted by color. One can see that 
the subproblem (a) has more complex structure as compared to the subproblem 
(b). At the same time, computing the values of the function $f(y_1,y_2)$ is 
more computation-costly than of $g(y_3,y_4)$.

@@

Let us conduct the next experiment employing $p$ nodes each including three 
graphic accelerators. Thus, at $p=32$ total 96 GPU accelerators were employed 
each including 2688 CUDA cores. In table 2, the execution time of the 
algorithm is presented and in table 3 -- the speedup with respect to the run 
on a single node.

@@

Computational experiments were carried out on a high-performance cluster of 
Lobachevsky State University of Nizhni Novgorod. The cluster node included 
two Intel Sandy Bridge E5-2660 2.2 GHz CPUs, 64 Gb RAM, and three NVIDIA 
Kepler K20Õ GPUs (2688 CUDA cores, 6 Gb GDDR5). 
The results of experiments demonstrate that the paralle index algorithm combined with the nested optimization scheme provides a good speedup on the considered problem class.

@@

\section*{Acknowledgments}
This study was supported by the Russian Science Foundation, project No 16-11-
10150.

@@

>>>>

\address{National Research Lobachevsky State University of Nizhny Novgorod, Gagarin avenue 23, 
Nizhny Novgorod, Russia, 603950}

@@

\section{Introduction}
In the present paper, the constrained multiextremal optimization problems and 
the parallel methods for solving these ones are considered. The fact that the 
global extremum is an integral characteristic of a problem is an important 
feature of the multiextremal problems. Thus, its finding is related to the 
construction of a coverage of the search domain and the computing the 
optimized function values at all points of this coverage. The dimensionality 
affects the difficulty of solving the problems of considered class crucially: 
the computational costs grow with increasing this one exponentially, 
therefore, for solving such problems, the methods are required, which 
generate a nonuniform grid in the search domain, denser near the 
global minimizer and more sparse far away from this one. In the 
present paper, the results of scalability investigation of the parallel 
index global optimization algorithm developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Strongin2000,Strongin2013} and implemented in the Globalizer 
system \cite{Globalizer} are presented.

@@

In the algorithms considered in the present paper the objective function and constraints are assumed to be Lipschitzian that is typical for many other approaches to the development of the parallel optimization algorithms (see, for example, \cite{Jones,Zilinskas,Evtushenko}). This assumption is natural for many applied problems since the relative variations of the functions generally cannot exceed a certain threshold determined by the bounded energy of changes occurring in the system under study. Also, a standard assumption is that the execution of a single trial (the computing of the values of the objective function and constraints at a point in the search domain) is a time-consuming operation since it 
utilized the results of numerical modeling.

An assumption on different costs of computing the problem functions subject to 
various components of the vector of parameters is a novel element investigated in the present work. 
It is assumed that it is possible to select the ``difficult'' and ``easy'' parts 
of the functions connected with each other. Then, the solving of the 
``difficult'' part of the problem (for which the execution of time-consuming algorithms is required to conduct the trials) can be performed on CPU using an efficient index global optimization algorithm. The ``easy'' part 
of the problem (the execution of which doesn't require the complex algorithms 
and can be transferred to an accelerator easily) can be solved on GPU using 
the uniform grid technique. Obviously, in this case a major portion of the trials 
will be carried out on GPU, and the minor one will be carried out on CPU. However, 
because of the difference in the difficulties of the subproblems solved on 
different units, a speedup of the algorithm as a whole can be expected. 
The proposed approach has been implemented in Globalizer parallel software 
system for solving the global optimization problem developed in Lobachevsky 
State University of Nizhny Novgorod.

@@

The use of Peano curve $y(x)$ 
\begin{equation}
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq 
N\right\}=\left\{y(x):0\leq x \leq 1 \right\}
\end{equation}
unambiguously mapping the interval of real axis $[0,1]$ onto a $N$-dimensional cube is the first of the dimensionality reduction methods considered.  To implement this method of dimensionality reduction a 
numerically constructed curve (\textit{evolvent}) is used. The evolvent is 
$2^{-m}$ accurate approximation of the theoretical Peano curve in $L_\infty$ 
metric, where $m$ is an evolvent construction parameter. Problems of 
numerical construction of the evolvents and the corresponding theory are 
considered in detail in \cite{Strongin2000}.

@@

Nested optimization scheme is based on relation (see \cite{Strongin2013})
\begin{equation}\label{nested}
\min_{y \in D}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq 
m\right\}}= \min_{u_1\in D_1}\min_{u_2\in D_2}...\min_{u_M\in D_M 
}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq m\right\}},
\end{equation}
which allows replacing the solving of multidimensional problem 
(\ref{problem}) by solving a family of subproblems related to each other 
recursively.
Here we consider vector $y$ as a vector of block variables
\begin{equation}
y=(y_1,y_2,...,y_N)=(u_1,u_2,...,u_M),
\end{equation}
where the $i$-th block variable $u_i$ is a vector of $N_i$ components of 
vector $y$, taken serially, i.e. $u_1=(y_1,y_2,...,y_{N_1})$, 
$u_2=(y_{N_1+1},y_{N_1+2},...,y_{N_1+N_2})$,..., $u_M=(y_{N-N_M+1},y_{N-
N_M+2},...,y_{N})$, at that $N_1+N_2+...+N_M=N$. 
The subdomains $D_i, 1 \leq i \leq M$, are projections of initial search 
domain $D$ onto the subspaces corresponding to the variables $u_i, 1 \leq i 
\leq M$.

@@

The organization of parallel computing with the use of the recursive 
optimization scheme has been considered in details for the shared/distributed 
memory as well as for the accelerators in \cite{Sysoyev,BarkalovLebedev}. However, in these works the 
problems were considered, in which the time of the trial execution didn't 
depend on the trial point. Here, the problem is considered assuming that in 
the function $\varphi(y_1,...,y_N)$ it is possible to select more difficult 
part $f(y_1,...,y_s)$ (depending on a part of the parameters only) and a simpler 
part $g(y_1,...,y_N)$ (depending on all problem parameters), for 
example, $\varphi(y)= f(y)g(y)$.

The difficult part of the function implies performing some  
computations related to the numerical simulation, which can be performed on the 
CPUs only. The simple part doesn't imply the complex computations and can be 
computed on an accelerator, for example, GPU. For solving a problem with such 
a structure, one can apply the parallel recursive optimization scheme with 
the use CPU at the upper nesting level and GPU at the lower one.

@@

The recursive scheme of solving the global optimization problems has been 
implemented in the Globzlizer solver developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Globalizer}. The global search methods and various 
dimensionality reduction schemes make the algorithmic base for the 
Globalizer. The numerical experiments, the results of which are presented in 
\cite{BarkalovGergelLebedev,BarkalovGergel} demonstrate these methods, at least, are not worse than the well known 
methods applied for similar purposes, and even overcome these ones with 
respect to some parameters.

@@

The level lines of the two-dimensional subproblems based on the functions 
$f(y_1,y_2)$ and $g(y_3,y_4)$ are shown in figure~1. The feasible domains are highlighted by color. One can see that 
the subproblem (a) has more complex structure as compared to the subproblem 
(b). At the same time, computing the values of the function $f(y_1,y_2)$ is 
more time-consuming than of $g(y_3,y_4)$.

@@

Let us conduct the next experiment employing $p$ nodes each including three 
graphic accelerators. Thus, at $p=32$ total 96 GPU accelerators were employed 
each including 2688 CUDA cores. In tables 2 and 3 the average time and the speedup with respect to the run 
on a single node are presented.

@@

Computational experiments were carried out on a high-performance cluster of 
Lobachevsky State University of Nizhny Novgorod. The cluster node included 
two Intel Sandy Bridge E5-2660 2.2 GHz CPUs, 64 Gb RAM, and three NVIDIA 
Kepler K20Õ GPUs (2688 CUDA cores, 6 Gb GDDR5). 
The results of experiments demonstrate that the parallel index algorithm combined with the nested optimization scheme provides a good speedup on the considered problem class.

@@

\section*{Acknowledgments}
This study was supported by the Russian Science Foundation, project No 16-11-10150.

@@

\bibitem{Globalizer}
Sysoyev, A.V. Globalizer -- A parallel software system for solving global optimization problems / A.V. Sysoyev, K.A. Barkalov, V.V. Sovrasov, I.G. Lebedev, V.P. Gergel // Lecture Notes in Computer Science. --- 2017. --- Vol. 10421. --- P. 492--499.
